---
title: "NASGLP - predicting mineralogy"
output: html_document
date: "2026-03-25"
---


Packages Needed

```{r Packages needed, message=FALSE, warning =FALSE}

# spatial packages 
library(terra)
library(sp)
library(sf)
library(rnaturalearth)
library(tidyterra)
library(sf)
library(spatstat)
library(mapview)

# general vis and data tidying 
library(dplyr)
library(ggplot2)
library(tidyr)
library(cowplot)
library(viridis)
library(scales)
library(CHNOSZ)
library(RNetCDF)
library(stringr)

#ML packages 
library(mlr3)
library(mlr)
library(ParamHelpers)
```

First let's read in NASGLP data. Only 2007 data has QXRD data, so we are leaving the 2004 pilot data out. 

```{r,warning =FALSE}
NASGLP_USA_2007 <- read.csv("C:/Users/lmw63/OneDrive - Yale University/Carbon-DOE_Synthesis/NASGLP_ML/NASGLP_ML_Workflow/Appendix_3a_Ahorizon_18Sept2013_NASGLP_2007.csv",skip = 12)   
# remove extra top row 
NASGLP_USA_2007 <- NASGLP_USA_2007[-1, , drop = FALSE]
# add a dataset ID
NASGLP_USA_2007$DatasetID <- "NASGLP_USA_2007"


# For the NAGLP_USA_2007 dataset, need to: 
  # 1) string replace all the NAs and Ns entries, and make columns of interest numerics 
  # 2) Filter for measurements with no carbonates detected 
  # 3) Select columns that we want, which include: ID, State/Province ID, Lat, Long, Depth, Total C, Organic C, and Dataset ID 
  # 4) Name these columnds the same thing as the NASGLP USA 2004 dataset 


NASGLP_USA_2007$A_C_tot <- str_replace_all(NASGLP_USA_2007$A_C_Tot,c("N.S."="NA", "N.D."=0.02)) 
NASGLP_USA_2007$A_C_org <- str_replace_all(NASGLP_USA_2007$A_C_Org, c("N.S."="NA", "N.D."=0)) 
NASGLP_USA_2007$A_C_Inorg <- str_replace_all(NASGLP_USA_2007$A_C_Inorg,c("N.S."="NA", "N.D."=0)) 
# ND based on minimum detection limit based on metadata
NASGLP_USA_2007$A_Al <- str_replace_all(NASGLP_USA_2007$A_Al,c("N.S."="NA", "N.D."=0.01)) 
NASGLP_USA_2007$A_Ca <- str_replace_all(NASGLP_USA_2007$A_Ca,c("N.S."="NA", "N.D."=0.01)) 
NASGLP_USA_2007$A_Fe <- str_replace_all(NASGLP_USA_2007$A_Fe,c("N.S."="NA", "N.D."=0.01)) 
NASGLP_USA_2007$A_K <- str_replace_all(NASGLP_USA_2007$A_K,c("N.S."="NA", "N.D."=0.01)) 
NASGLP_USA_2007$A_Mg <- str_replace_all(NASGLP_USA_2007$A_Mg,c("N.S."="NA", "N.D."=0.01)) 
NASGLP_USA_2007$A_Na <- str_replace_all(NASGLP_USA_2007$A_Na,c("N.S."="NA", "N.D."=0.01)) 

NASGLP_USA_2007$A_Tot_10A <- str_replace_all(NASGLP_USA_2007$A_Tot_10A,c("N.S."="NA", "N.D."=0.02)) 
NASGLP_USA_2007$A_Tot_14A <- str_replace_all(NASGLP_USA_2007$A_Tot_14A,c("N.S."="NA", "N.D."=0.02)) 
NASGLP_USA_2007$A_Kaolinit <- str_replace_all(NASGLP_USA_2007$A_Kaolinit,c("N.S."="NA", "N.D."=0.02)) 

NASGLP_USA_2007$A_C_Inorg <-as.numeric(NASGLP_USA_2007$A_C_Inorg)
NASGLP_USA_2007$A_C_tot <- as.numeric(NASGLP_USA_2007$A_C_Tot) 
NASGLP_USA_2007$A_C_org <- as.numeric(NASGLP_USA_2007$A_C_Org) 

NASGLP_USA_2007$A_Al <- as.numeric(NASGLP_USA_2007$A_Al)
NASGLP_USA_2007$A_Ca <- as.numeric(NASGLP_USA_2007$A_Ca)
NASGLP_USA_2007$A_Fe <- as.numeric(NASGLP_USA_2007$A_Fe)
NASGLP_USA_2007$A_K <- as.numeric(NASGLP_USA_2007$A_K)
NASGLP_USA_2007$A_Mg <- as.numeric(NASGLP_USA_2007$A_Mg)
NASGLP_USA_2007$A_Na <- as.numeric(NASGLP_USA_2007$A_Na)
NASGLP_USA_2007$A_Tot_10A <- as.numeric(NASGLP_USA_2007$A_Tot_10A)
NASGLP_USA_2007$A_Tot_14A <- as.numeric(NASGLP_USA_2007$A_Tot_14A)
NASGLP_USA_2007$A_Kaolinit <- as.numeric(NASGLP_USA_2007$A_Kaolinit)

```
Our goal is to see if we can predict phyllosilicate composition with our geochemical elements of interest. The phyllosilicates of interest are: A_Tot_10A (Micas), A_Tot_14A (Expandable clays), A_Kaolinit (Our favorite 1:1)

And our geochemical elements of interest are: A_Al,A_Ca, A_Fe, A_K, A_Mg, A_Na

Going to select our response annd predictor variables and extract coordinates first. 

I am going to create 3 models - one to predict A_tot_10A, one to predict A_Tot_14A, and one to predict A_kaolinite. 

```{r}

NASGLP_clean <- NASGLP_USA_2007 %>%
  drop_na(A_Tot_10A,A_Tot_14A,A_Kaolinit,A_Al,A_Ca,A_Fe,A_K,A_Mg,A_Na)

coords_data <- NASGLP_clean[,c("Longitude","Latitude")]

data_10A <- NASGLP_clean %>%
  select(A_Al,A_Ca,A_Fe,A_K,A_Mg,A_Na,A_Tot_10A)

data_14A  <- NASGLP_clean %>%
  select(A_Al,A_Ca,A_Fe,A_K,A_Mg,A_Na,A_Tot_14A)

data_kaolin  <- NASGLP_clean %>%
  select(A_Al,A_Ca,A_Fe,A_K,A_Mg,A_Na,A_Kaolinit)




```

We are going to create out regression tasks 

```{r}

task_10A <- makeRegrTask(data=data_10A,target = "A_Tot_10A", coordinates = coords_data)

task_14A <- makeRegrTask(data=data_14A,target = "A_Tot_14A", coordinates = coords_data)

task_kaolin <- makeRegrTask(data=data_kaolin,target = "A_Kaolinit", coordinates = coords_data)



```

Create learner 

```{r}

lrn_rf <- makeLearner(cl="regr.ranger",predict.type = "response")

```


Specify resampling


```{r}

tune_level <- makeResampleDesc("SpCV",iters=5)
```

Specify number of random search

```{r}

ctrl <- makeTuneControlRandom(maxit = 50L)

```

Search for hyperparamters done 50x in each spatially disjoint partition, which takes spatial bias into account 

```{r}

ps_10A <- ParamHelpers::makeParamSet(
makeIntegerParam("mtry", lower = 1, upper = ncol(data_10A) - 1),
makeNumericParam("sample.fraction", lower = 0.2, upper = 0.9),
makeIntegerParam("min.node.size", lower = 1, upper = 10))

ps_14A <- ParamHelpers::makeParamSet(
makeIntegerParam("mtry", lower = 1, upper = ncol(data_14A) - 1),
makeNumericParam("sample.fraction", lower = 0.2, upper = 0.9),
makeIntegerParam("min.node.size", lower = 1, upper = 10))

ps_kaolinit <- ParamHelpers::makeParamSet(
makeIntegerParam("mtry", lower = 1, upper = ncol(data_kaolin) - 1),
makeNumericParam("sample.fraction", lower = 0.2, upper = 0.9),
makeIntegerParam("min.node.size", lower = 1, upper = 10))


```

Tuning for 10A prediction model:

```{r}

tune_10A <- tuneParams(learner=lrn_rf,
                       task= task_10A,
                       resampling = tune_level,
                       par.set=ps_10A,
                       control=ctrl,
                       measures=mlr::rmse)

```


```{r}

set.seed(42)
lrn_rf_10A <- setHyperPars(makeLearner("regr.ranger",predict.type = "response"),
                           par.vals = tune_10A$x)

set.seed(42)
model_rf_10A <- train(lrn_rf_10A,task_10A)

getLearnerModel(model_rf_10A)

```



Tuning for 14A prediction model: 

```{r}

tune_14A <- tuneParams(learner=lrn_rf,
                       task= task_14A,
                       resampling = tune_level,
                       par.set=ps_14A,
                       control=ctrl,
                       measures=mlr::rmse)


```



```{r}

set.seed(42)
lrn_rf_14A <- setHyperPars(makeLearner("regr.ranger",predict.type = "response"),
                           par.vals = tune_14A$x)

set.seed(42)
model_rf_14A <- train(lrn_rf_14A,task_14A)

getLearnerModel(model_rf_14A)


```

Tuning for kaolinite prediction model: 

```{r}

tune_kaolin <- tuneParams(learner=lrn_rf,
                       task= task_kaolin,
                       resampling = tune_level,
                       par.set=ps_kaolinit,
                       control=ctrl,
                       measures=mlr::rmse)

```



```{r}



set.seed(42)
lrn_rf_kaolin <- setHyperPars(makeLearner("regr.ranger",predict.type = "response"),
                           par.vals = tune_kaolin$x)

set.seed(42)
model_rf_kaolin <- train(lrn_rf_kaolin,task_kaolin)

getLearnerModel(model_rf_kaolin)


```



